---
title: "Handout 09: Strings in R"
date: Taylor Arnold
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
---


```{r global_options, include=FALSE}
library(tufte)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(tidy = FALSE)

library(tidyverse)
library(smodels)
library(forcats)
library(ggrepel)
library(stringr)

theme_set(theme_minimal())
msleep <- na.omit(msleep)
```

\justify

# Basic string manipulation

This handout covers the main aspects of working with raw
strings in R using the **stringr** package. This is part
of the tidyverse suite, but not loaded by default. To load
the package call:
```{r}
library(stringr)
```
The main advantages of this package over this package compared
to those in base-R are:

* consistent syntax - the string you are operating on is always
the first element and functions all start with `str_`
* great support for non-latin character sets and proper UTF-8
handling
* in some cases much faster than alternatives

We will work with two datasets that come pre-installed with
**stringr**, a list of common English tokens named `words`
and a list of short sentences named `sentences`. We will
wrap these up as data frames in order to make them usable
by the **dplyr** verbs we have been learning:

```{r}
df_words <- data_frame(words = words)
df_sent <- data_frame(sent = sentences)
```

The first function we will look at is `str_sub` that takes
a substring of each input by position; for example the following
finds the first three characters of every string in the data
set of words:

```{r}
mutate(df_words, chars = str_sub(words, 1, 3))
```

Notice that R silently ignores the fact that the first word
that has only one letter (it is returned as-is).
We can use negative values to begin at the end of the string
(-1 is the last character, -2 the second to last and so on).
So the last two characters can be grabbed with this:

```{r}
mutate(df_words, chars = str_sub(words, -2, -1))
```

The function `str_length` describes how many characters are
in a string:

```{r}
mutate(df_words, num_char = str_length(words))
```

And the functions `str_to_lower` and `str_to_upper` do exactly
as they describe:

```{r}
mutate(df_words, up = str_to_upper(words), down = str_to_lower(words))
```

We even have `str_to_title` to convert to title case:

```{r}
str_to_title("The birch canoe slid on the smooth planks.")
```

# Matching strings

We now move on to some more complex string functions that
require matching some type of pattern, starting with very
simply examples. Take the function `str_view_all` that
visually displays all occurrences of a pattern in RStudio:

```{r, eval = FALSE}
str_view_all(df_sent$sent, "is")
```

Notice that you can select to *show in new window* the results
in the View pane of RStudio to see all of the results. Notice
that this pattern reveals the word "is" but also other
occurrences where this string occurs within a word.

Another function that finds patterns is the function
`str_detect`, which returns either `TRUE` or `FALSE`
for whether an element has a string withing in. We
can use this conjunction with the `filter` command
to find examples with a particular string in it:

```{r}
filter(df_sent, str_detect(sent, "hand"))
```

Similarly `str_count` tells us how often a sentence
uses a particular string. For instance, how many times
are the digraphs "th", "ch", and "sh" used in each
sentence:

```{r}
mutate(df_sent, th = str_count(sent, "th"),
                sh = str_count(sent, "sh"),
                ch = str_count(sent, "ch"),
                sent = str_sub(sent, 1, 20))

```

I took a substring of the first column to make it fit on
the page.

The function `str_replace_all` replaces one pattern with
another. Perhaps we want to replace all of those borning
"e"'s with "ë":

```{r}
mutate(df_sent, sent = str_replace_all(sent, "e", "ë"))
```

The function `str_replace` without the "all" only replaces
the first occurrence in each string.

# Matching patterns

Trying to use the previous functions with a fixed string
can be useful, but the true strength of these functions
come from their ability to accept a pattern known as a
regular expression. We don't have time to cover these in
great detail, but will show a few important examples. The
first example we will us is the "." symbol which matches
any character. So, for instance this finds any time that
we have the letters "w" and "s" separated by any third
character:

```{r}
filter(df_sent, str_detect(sent, "w.s"))
```

Can you find where this occurs in each line? Two other
special characters are "^" and "$", called *anchors*.
The first matches  the start of a sentence and the
second matches the end of a sentence. So, which words
end with the letter "w"?

```{r}
filter(df_words, str_detect(words, "w$"))
```

Or start with "sh"?

```{r}
filter(df_words, str_detect(words, "^sh"))
```

How would we actually match a literal period, dollar sign,
or other special character? The answer is to put two slashes
before it; so this is how to replace all of the periods
in the sentences into exclamation marks:

```{r}
mutate(df_sent, sent = str_replace(sent, "\\.", "!"))
```

There is on other **string** function we did not mention
earlier: `str_extract`. Given a pattern it returns the
string that matches it. This is not very useful without
regular expression but with them is an invaluable tool.
For example, what characters follow the pattern "th"?

```{r}
temp <- mutate(df_sent, triple = str_extract(sent, "th."))
table(temp$triple)
```


There are many other more complex regular expressions. For
example, this one is very useful:

```{r, eval = FALSE}
str_replace(html, "<[^>]+>", " ")
```

If `html` is a string, this will replace all of the characters
in html tags with a single space. We will use that in our lab
today.
