---
title: "Handout 05: Data Manipulation"
date: Taylor Arnold
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
---


```{r global_options, include=FALSE}
library(tufte)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(tidy = FALSE)

library(tidyverse)
library(smodels)
library(forcats)
library(ggrepel)

theme_set(theme_minimal())
```

\justify

\newthought{So far we have primarily} been working with a single
dataset as it was directly given to us. In many cases it is
advantageous to produce new datasets from our original one. This
can be as simple as selecting a subset of the original columns
or rows, or as involved as changing the *level of analysis* of
the entire dataset. This handout will introduce several functions
called *verbs* that assist in manipulating datasets. These all come
from the package **dplyr**, contained in **tidymodels**.

In this handout, as with the others so far, I will use the `msleep`
dataset in order to show various numerical summaries. To read it in
use the following:

```{r}
data(msleep)
```

# A grammar for data manipulation

The *verbs* in **dplyr** take either one or
two existing datasets and return a new dataset. There are roughly 30
different verbs, of which we'll use only about 4 in this course:^[
Actually, `group_summarize` is a verb that I wrote specifically for this
course and comes from the **smodels** package. It combines several **dplyr**
verbs in a particularly useful way.]

* `filter`
* `mutate`
* `group_summarize`
* `left_join`

These verbs on their own are relatively straightforward, but can be tied
together to produce suprisingly complex new datasets.

## Filtering rows

The filter function takes a dataset followed by a logical statements.
It returns a dataset that has any rows in the input data that match the
filtering statements. For example, the following returns a dataset for
all mammals that are awake more than 20 hours per day:

```{r echo = TRUE}
filter(msleep, awake > 20)
```

It is very important to notice that the original `msleep` dataset has not been
altered here. If we want to actually work with the filtered data,
we need to save it using the assignment operator `<-` and given it a name. To
pull out just the *sleepy* mammals, we could do the following:

```{r echo = TRUE, eval = FALSE}
sleepy <- filter(msleep, awake < 6)
```

You should see now that a new dataset appears in your workspace named `sleepy`.
It is possible to work with this new dataset in all of the ways we have plotted and
(now) filtered the original data.

## Constructing new variables

The `mutate` function preserves all rows of the original dataset, unlike `filter`,
but adds a new variables. For example, to add hours asleep into the dataset
we can do this:

```{r echo = TRUE}
msleep <- mutate(msleep, asleep = 24 - awake)
```

Here, I used the assignment command to save the result back into the `msleep`
dataset. If you look in your environment window, you'll still see just a single
version of `msleep` but this one will have one extra variables. This is a relatively
safe pratice with the mutate function, as we are simply adding information, but
should be generally avoided when using filter.^[It is possible to use mutate to
redefine an existing variable by giving mutate a variable name that already exists.
Be careful of this, particularly if you are overwritting the original dataset.]

## Summarizing data

The `group_summarize` function is, in my opinion, the most complex verb that
we will use this semester. If we use the function on a dataset without any
other options it gives the mean, median, standard deviation, and sum for
every numeric  variable in the dataset. An overall count is also included.
Let's apply it to `msleep`:

```{r echo = TRUE, eval = TRUE}
group_summarize(msleep)
```

These variables could have easily been computed by calling the respective
functions indvidually in R. The group summarize function becomes more
interesting when we pass it a second input giving a variable to group by.
For example, here is the summary *grouped by vore*

```{r echo = TRUE, eval = TRUE}
group_summarize(msleep, vore)
```

Notice that the result now provides these summaries for each group. It
is possible to summarize by multiple groups at once, which produces
summaries for each unique combination of the those variables. For instance,
we could summarize by both genus and vore:

```{r echo = TRUE, eval = TRUE}
group_summarize(msleep, vore, genus)
```

And there is now a row for each combination of `vore` and `genus`.

# Combine datasets

The final verb that we will use this semester is also the only two-table
verb that we will need. It will be used to combine a dataset with metadata
about one or more of its variables. To illustrate, let's make a small
dataframe that contains the full name for the short-hand abbreviations
given in the variable `vore`:

```{r echo = TRUE}
meta <- data_frame(vore = c("carni", "omni", "herbi", "insecti"),
                   full_name = c("carnivore", "omnivore", "herbivore",
                                 "insectivore"))
meta
```

To combine these with the original dataset, we use the `left_join` function,
giving the larger dataset first:

```{r echo = TRUE}
msleep <- left_join(msleep, meta)
msleep
```

Notice that there is a new variable `full_name` that now displays the
full name for each `vore` type.



