---
title: "Handout 02: Graphics in R"
date: Taylor Arnold
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
---


```{r global_options, include=FALSE}
library(dplyr)
library(ggplot2)
library(tufte)
knitr::opts_chunk$set(results = "hide")
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(tidy = FALSE)

library(ggplot2)
library(ggrepel)
library(dplyr)
library(readr)
library(MASS)

theme_set(theme_minimal())
```

\justify

\newthought{One of the most important} tasks in data analysis is
visualizing the data. This is important from the first step of
understanding a new dataset all the way through producing graphics
for a final presentation. We will use a package called **ggplot** in
order to produce visualizations this semester. The package provides
a particular plotting paradigm that has a solid theoretical
underpinning.

In this handout, I will use the `msleep` dataset in order to show
various plots. This data is actually includes with the **ggplot2**
package, and can be loaded as follows:

```{r}
data(msleep)
```

The data contains information about the sleep patterns of 83
different mammals.^[V. M.
Savage and G. B. West. A quantitative, theoretical framework for
understanding mammalian sleep. Proceedings of the National Academy
of Sciences, 104 (3):1051-1056, 2007.]

# Basic plotting

## One numeric variable

The default plot for a single numeric variable is a *histogram*. To
get the plot simply use the following:

```{r}
qplot(awake, data = msleep)
```

As described on Wikipedia, which I suggest looking at for more information
if you are still unclear about the definition of a histogram:

> A histogram is a graphical representation of the distribution of numerical
> data. It is an estimate of the probability distribution of a continuous
> variable (quantitative variable) and was first introduced by Karl Pearson.
> To construct a histogram, the first step is to "bin" the range of values
> - that is, divide the entire range of values into a series of intervals -
> and then count how many values fall into each interval. The bins are usually
> specified as consecutive, non-overlapping intervals of a variable. The bins
> (intervals) must be adjacent, and are often (but are not required to be)
> of equal size.

By default the `qplot` function uses 30 bins, but this can be changed to
any number you would like. Notice how changing the bins to include only 5
changes the plot:

```{r}
qplot(awake, data = msleep, bins = 5)
```

This makes the bins even more prominent. To test your knowledge, how many
mammals are awake less than (about) 11 hours per day? You should be able
to get this quickly from the plot.

## One discrete variable

The default plot for a categorical variable is a *bar plot*. The code to
generate it is exactly the same as for the histogram:

```{r}
qplot(vore, data = msleep)
```

A bar plot is very similar to a histogram except that there is no need to
bin the data into categories.

## Two continuous variable

To plot two continuous variables, we simply supply both variables to the
`qplot` function:

```{r}
qplot(awake, sleep_rem, data = msleep)
```

The result is a *scatter plot*, which you have almost certainly seen at some
point. Each data point is represented by a point on the plot.

## One continuous and one discrete variable

The default plot for two variables where one is discrete does not have a
special name, but can be produced using a similar syntax:

```{r}
qplot(awake, vore, data = msleep)
```

It resembles a scatter plot, except that one axis is a discrete categorical
variable.

# Aesthetics

There are a number of graphical properties that we can manipulate on a plot,
usually in the context of a scatter plot. The basic syntax behind these are
discussed below. For each, there is a way of changing the properties either
uniformly for every point or based on a third variable.

## Color

To change the color of points based on a third variable, simply supply the
color parameter to the function `qplot` with the variable name:

```{r}
qplot(awake, sleep_rem, data = msleep, color = vore)
```

Notice that a legend appears to describe what each color represents. We can
also use a continuous variable to specify color with the same syntax:

```{r}
qplot(awake, sleep_rem, data = msleep, color = sleep_cycle)
```

If we instead want to change every color uniformly, we instead wrap the name
of the color in quotes and the function `I`:

```{r}
qplot(awake, sleep_rem, data = msleep, color = I("salmon"))
```

There are a lot of colors to choose from in R; to see all of the options type
`colors()` in the console.

## Size

It is also possible to change the size of the points in a similar fashion. Here
is does not make sense to use a categorical variable, so we will stick to numeric
variables:

```{r}
qplot(awake, sleep_rem, data = msleep, size = sleep_cycle)
```

This kind of scatter plot is called a *bubble chart*; we saw an example of
this in Hans Roslin's video. It is also possible to change the size uniformly
for all points, using the `I` function once again:

```{r}
qplot(awake, sleep_rem, data = msleep, size = I(0.2))
```

For reference, the default size is 1.

## Alpha

Finally, we can also change the opacity of the points. That is, to what extent
are the points see-through. The parameter that controls this is called
`alpha`; when set to 1 (the default) the point is not at all see-through and
when set to 0 it is invisible. Notice what happens when I set it to $0.2$:

```{r}
qplot(awake, sleep_rem, data = msleep, alpha = I(0.2))
```

It is possible to set the alpha value to change with a third variable, though
this is rarely very useful.

# Adding layers

It is often useful to add additional plot tops over the top of the default
types supplied by `qplot`. To do this we simply add functions that start
with `geom_`. For example, a line plot (which actually does not make
any sense here) can be produce by:

```{r}
qplot(awake, sleep_rem, data = msleep) + geom_line()
```

It looks strange in part because of missing values in the dataset. Another
common alternative plot when we have both a continuous and a categorical
variable is called a *box plot*:

```{r}
qplot(vore, awake, data = msleep) + geom_boxplot()
```

The details of this plot will be described in a future handout. We can also
add a smoothing curve with:

```{r}
qplot(awake, sleep_rem, data = msleep) + geom_smooth()
```

If we want a best fit line, the option `method` can be set to `lm`:

```{r}
qplot(vore, awake, data = msleep) + geom_smooth(method = "lm")
```

These plots will be very helpful as we approach statistical modelling in
a few weeks.



