---
title: "Handout 03: Data Types"
date: Taylor Arnold
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
---


```{r global_options, include=FALSE}
library(tufte)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(tidy = FALSE)

library(tidyverse)
library(smodels)
library(forcats)
library(ggrepel)

theme_set(theme_minimal())
```

\justify

\newthought{We have already discussed} that variables in R
have different *data types*, most likely one of the following:

* `int` stands for integers.
* `dbl` stands for doubles, or real numbers.
* `chr` stands for character vectors, or strings.

And we can see these when printing out the data, such as with
the mammals sleep data:

```{r}
data(msleep)
msleep
```

The names and genus are characters whereas numbers such as `awake`
are doubles. This handout shows how to convert numeric variables
into categorical ones and way to manipulate categorical variables.
There is generally no straightforward way to convert a categorical
variable into a number, so we do not discuss that here.

# Numeric to Categorical

## With factor()

To convert every unique values of a numeric variable to a category
in a new categorical variable, we simply use the function `factor`.
For example, we can convert `sleep_rem` into categories. This changes,
amongst other things, the way color is used in a plot:

```{r}
qplot(vore, awake, data = msleep, color = factor(sleep_rem))
```

Notice that missing values became there own category. This is often
very useful.

## With cut() and bin()

Often we do not want each numeric variable to be its own category but
instead wish to group values that are close into a single category.
There are two related functions for doing this: `cut` and `bin`. Both
take the variable followed by the number of bins to use. The function
`cut` splits the range of the variable into equally spaced buckets
where as `bin` breaks the range up so that each bin has (roughly) the
same proportion of data points.

For most variables the difference between these two functions is small;
I usually use `cut` because the interval cutoffs print out nicer. However,
when a variable is fairly skewed, I find that `bin` works better. For
example, notice how different cutting and binning the `brainwt` variable
is:

```{r}
qplot(vore, awake, data = msleep, color = cut(brainwt, 5))
qplot(vore, awake, data = msleep, color = bin(brainwt, 5))
```

This variable is very skewed and the cutting algorithm puts almost all of
the mammals in the same category.


# Manipulating Categorical Data

Likewise, it is sometimes useful to convert one categorical variable into
another categorical variable by grouping or ordering the categories in a
different way.

By default categories are order alphabetically. The function `fct_inorder`
instead arranges them in their order of appearance. Compare, for example
the following:

```{r}
qplot(vore, data = msleep)
qplot(fct_infreq(vore), data = msleep)
```

The function `fct_lump` take a second argument that gives the maximum
number of categories allowed. The most frequent categories are included
and all other categories are lumped into the `other` category.

```{r}
qplot(vore, data = msleep)
qplot(fct_lump(vore, 2), data = msleep)
```

Finally, the function `fct_relevel` takes a second argument that gives
the category that should be first:

```{r}
qplot(fct_relevel(vore, "omni"), data = msleep)
```

This latter function may seem odd here but will be indispensable when used
with the modeling functions that will be introduced in a few weeks.


