---
title: "Handout 04: Advanced graphics"
date: Taylor Arnold
output:
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
---


```{r global_options, include=FALSE}
library(dplyr)
library(ggplot2)
library(tufte)
knitr::opts_chunk$set(results = "hide")
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(tidy = FALSE)

library(ggplot2)
library(ggrepel)
library(dplyr)
library(readr)
library(MASS)

theme_set(theme_minimal())
```

\justify

\newthought{We have already seen how to build fairly} complex
graphics within R. This handout fills in a few gaps that will
serve us well throughout the remainder of the semester.

In this handout, I will again use the `msleep` dataset in
my examples:

```{r}
data(msleep)
```

# Multiple Datasets

It is possible to plot multiple datasets in a single graphic. This
is particularly useful when combined with the `filter` function. For
example, we can highlight the insectivores by filtering them into a
seperate dataset:

```{r}
insecti <- filter(msleep, vore == "insecti")
```

And then ploting all of the mammals in black, but the insectivores
in red. This is accomplished by using the `geom_point` function and
giving it the name of the secondary dataset:

```{r fig.width=8, fig.height=8}
qplot(awake, sleep_rem, data = msleep) +
  geom_point(data = insecti, color = I("red"))
```

We can add multiple layers to the same plot. For example, the
`geom_text_repel` adds smart labels to the plot; using it we can label
all of the insectivores in our dataset:

```{r fig.width=8, fig.height=8}
qplot(awake, sleep_rem, label = name, alpha = I(0.5), data = msleep) +
  geom_point(data = insecti, color = I("red")) +
  geom_text_repel(data = insecti)
```

These types of plots become very useful for understanding how one subset
relates to a larger collection of points.

# Labels

The default labels on our plots are mostly fine for exploratory analysis.
When presenting your work in a paper or other venue it can, however, be
useful to modify these with more descriptive terms. To do so, we simply
add the functions `xlab`, `ylab`, and `ggtitle` to the plot. For example:

```{r fig.width=8, fig.height=8}
qplot(awake, sleep_rem, label = name, alpha = I(0.5), data = msleep) +
  geom_point(data = insecti, color = I("red")) +
  geom_text_repel(data = insecti) +
  xlab("Average hours awake in 24 hour time period") +
  ylab("Average hours spent in REM sleep in 24 hour period") +
  ggtitle("Insectivore sleep patterns relative to other mammals")
```

You may of course add only those elements you wish to choose; those not
specified will be left at their default values.



